Switch equivalence and weak prime domains for left-linear DPO rewriting systems

### 2025-03-07 full_version/leftlinear.tex
minima

\begin{example}	\label{ex:seq-ind}
	Consider the rewriting system  $(\cat{Graph}, \R)$ and the derivarion $\der{D}$ introduced in \Cref{ex:1}.	Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
	independent via the arrows $R_1 \to D_2$, $L_2 \to D_1$
	mapping nodes according to their numbering.

@@

\Cref{ex:diff1} show that sequential independence it does not suffice, in general, to guarantee the existence of a switch.  Inspired by the notion of of \emph{canonical filler}
\cite{heindel2009category}, we are now going to give some conditions guaranteeing the existence of a switch. We start proving some technical properties of independence pairs.

@@

To see that they are one the inverse of the other it is then enough to compute.
		\[\begin{split}
		t_0\circ v\circ w &=\phi_{Q_0}\circ q_0\circ w\\&=\phi_{Q_0}\circ \phi^{-1}_{Q_0} \circ t_0\\&=t_0\\q_0\circ w\circ v&=\phi^{-1}_{Q_0}	\circ t_0\circ v\\&=\phi^{-1}_{Q_0} \circ \phi_{Q_0}\circ q_0\\&=q_0 	\end{split}\qquad \begin{split}
		t_1\circ v\circ w &=\phi_{Q_1}\circ q_1\circ w\\&=\phi_{Q_1}\circ \phi^{-1}_{Q_1} \circ t_1\\&=t_1\\q_1\circ w\circ v&=\phi^{-1}_{Q_1}	\circ t_1\circ v\\&=\phi^{-1}_{Q_1} \circ \phi_{Q_1}\circ q_1\\&=q_1
		\end{split}\]

@@

 $G_1$ is obtained as the pushout of $k_0$ along $r_0$m thus get the wanted arrow $\kappa\colon G_1\to \tpro{E}$ it is now enough to compute:
	\begin{align*}
		\iota_{\der{E}, G_0}\circ f_{0}\circ  k_{0}&= \iota_{\der{E}, G_0}\circ m_{0} \circ l_{ 0}\\&=\iota_{\der{E}, G_0}\circ  f'_{0} \circ i'_1 \circ l_0\\&=\iota_{\der{E}, D'_0}\circ i'_1 \circ l_0\\&=\iota_{\der{E}, G'_1} \circ g'_0\circ m'_1 \circ l_0\\&=\iota_{\der{E}, G'_1}\circ m'_1\circ l_0	\\&=\iota_{\der{E}, G'_1} \circ f'_1\circ k'_1\\&=\iota_{\der{E}, D'_1}\circ k'_1\\&=\iota_{\der{E}, K'_0}\\&= \iota_{\der{E}, R_0} \circ r_0
	\end{align*}

@@

\begin{lemma}
Let $(P, \{p_0, p_1\})$ be the pullback of $f_1$ along $g_1$, then there exists $z_0\colon P\mto D'_0$ and $z_1\colon P\to D'_1$ making the following diagram commutative
\[\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){i_0}|(.7)\hole
	\ar[dr]^{h_0}&& L_1\ar@/^.8cm/[ddll]^(.2){i_1}
	\ar[dl]_{m_1}\\&K_0\ar[dr]^{k_0}\ar@{>->}[dl]_{l_0}
	\ar@/_.6cm/[ddrr]|(.36)\hole^(.65){u_0}\ar[ur]^{r_0}&& G_1 &&
	K_1
	\ar@/^.6cm/[ddll]|(.36)\hole_(.65){u_1}\ar[dl]_{k_1}\ar@{>->}[lu]_{l_1}
	\ar[dr]^{r_1}\\L_0
	\ar@/_.6cm/[ddrr]_(.2){i'_1}|(.31)\hole|(.81)\hole
	\ar[dr]^(.4){m_0}|(.61)\hole && D_0
	\ar@{>->}[dl]|(.4)\hole_(.65){f_0}\ar[ur]|(.5)\hole^(.7){g_0}&&D_1
	\ar[dr]|(.4)\hole^(.65){g_1}
	\ar@{>->}[ul]|(.5)\hole_(.65){f_1}&&R_1\ar@/^.6cm/[ddll]^(.2){i'_0}|(.31)\hole|(.81)\hole\ar[dl]_{h_1}|(.6)\hole\\&G_0
	&&P\ar@{.>}[dr]_{z_1}	\ar@{>.>}[dl]^{z_0}\ar[ur]^(.4){p_1}\ar@{>->}[ul]_(.4){p_0}&&G_2\\L_1	\ar@/^.6cm/[uurr]^(.2){i_1} \ar[ur]_(.35){m'_0}|(.61)\hole && D'_0	\ar@{>->}[ul]|(.4)\hole_(.65){f'_0}\ar[dr]|(.5)\hole_(.65){g'_0}	&&D'_1\ar[ur]|(.4)\hole^(.65){g'_1} \ar@{>->}[dl]|(.5)\hole^(.65){f'_1}	&& R_0 \ar[ul]^(.35){h'_1}|(.61)\hole\ar@/_.6cm/[uull]_(.2){i_0}\\ &K_1	\ar[ur]_{k'_0} \ar[dr]_{r_1}	\ar@{>->}[ul]^{l_1}\ar@/^.6cm/[uurr]_(.65){u_1}&&G'_1&& K_0\ar[ur]_{r_0} \ar[ul]^{k'_1} \ar@{>->}[dl]^{l_0} \ar@/_.6cm/[uull]^(.65){u_0}\\&& R_1	\ar[ur]_{h'_0}\ar@/^.8cm/[uurr]^(.2){i'_0}&& L_0 \ar[ul]^{m'_1} \ar@/_.8cm/[uull]_(.2){i'_1} |(.69)\hole }\] 
moreover, the following squares are all pushouts, moreover the first two are also pullbacks.
\[\xymatrix{P \ar@{>->}[r]^{z_0} \ar@{>->}[d]_{p_0}& D'_0 \ar@{>->}[d]^{f'_0} &  P \ar@{>->}[d]_{z_0} \ar[r]^{z_1} & D'_1 \ar@{>->}[d]^{f'_1} & P\ar[r]^{z_1} \ar[d]_{p_1}&D'_1 \ar[d]^{g'_1}\\ D_0 \ar@{>->}[r]_{f_0} & G_0 & D'_0 \ar[r]_{g'_0}& G'_1 & D_1 \ar[r]_{g_1}& G_2}\]
\end{lemma}
\begin{proof}
	Define $z_0$ and $z_1$ to be, respectively, $t_0\circ v$ and $t_1\circ v$. To prove the commutativity of the diagram in the thesis, first of all we can notice that
	\begin{align*}
		g'_0\circ z_0\\&= g'_0\circ t_0\circ v\\&=f'_1\circ t_1\circ v\\&=f'_1\circ z_1
	\end{align*}
	To prove the other identities, it is enough to compute and use the identities obtained in the previous point. On the one hand we have
	\[	\begin{split}f'_0\circ z_0&=f'_0\circ t_0\circ v\\&=f'_0\circ \phi_{Q_0}\circ q_0\\&=a_0\circ q_0\\&=f_0\circ p_0 
	\end{split} \qquad \begin{split}
		g'_1\circ z_1&=g'_1\circ t_1\circ v\\&=g'_1\circ \phi_{Q_1}\circ q_1\\&=\phi_{G_2}\circ b_1\circ q_1\\\phi_{G_2}\circ g_1\circ p_1\\&=
	\end{split}\]

@@

\begin{lemma}{lemma}{lemVTame}
	\label{bono}
	All root-preserving graphical rewriting systems are well-switching.
\end{lemma}

>>>>

\begin{example}	\label{ex:seq-ind}
	Consider the rewriting system  $(\cat{Graph}, \R)$ and the derivation $\der{D}$ introduced in \Cref{ex:1}.	Steps $\dder{D}_1$ and $\dder{D}_2$ are clearly sequential
	independent via the arrows $R_1 \to D_2$, $L_2 \to D_1$
	mapping nodes according to their numbering.

@@

\Cref{ex:diff1} show that sequential independence does not suffice, in general, to guarantee the existence of a switch.  Inspired by the notion of of \emph{canonical filler}
\cite{heindel2009category}, we are now going to give some conditions guaranteeing the existence of a switch. We start proving some technical properties of independence pairs.

@@

To see that they are one the inverse of the other is then enough to compute.
		\[\begin{split}
		t_0\circ v\circ w &=\phi_{Q_0}\circ q_0\circ w\\&=\phi_{Q_0}\circ \phi^{-1}_{Q_0} \circ t_0\\&=t_0\\q_0\circ w\circ v&=\phi^{-1}_{Q_0}	\circ t_0\circ v\\&=\phi^{-1}_{Q_0} \circ \phi_{Q_0}\circ q_0\\&=q_0 	\end{split}\qquad \begin{split}
		t_1\circ v\circ w &=\phi_{Q_1}\circ q_1\circ w\\&=\phi_{Q_1}\circ \phi^{-1}_{Q_1} \circ t_1\\&=t_1\\q_1\circ w\circ v&=\phi^{-1}_{Q_1}	\circ t_1\circ v\\&=\phi^{-1}_{Q_1} \circ \phi_{Q_1}\circ q_1\\&=q_1
		\end{split}\]

@@

 $G_1$ is obtained as the pushout of $k_0$ along $r_0$, thus to get the wanted arrow $\kappa\colon G_1\to \tpro{E}$ is now enough to compute:
	\begin{align*}
		\iota_{\der{E}, G_0}\circ f_{0}\circ  k_{0}&= \iota_{\der{E}, G_0}\circ m_{0} \circ l_{ 0}\\&=\iota_{\der{E}, G_0}\circ  f'_{0} \circ i'_1 \circ l_0\\&=\iota_{\der{E}, D'_0}\circ i'_1 \circ l_0\\&=\iota_{\der{E}, G'_1} \circ g'_0\circ m'_1 \circ l_0\\&=\iota_{\der{E}, G'_1}\circ m'_1\circ l_0	\\&=\iota_{\der{E}, G'_1} \circ f'_1\circ k'_1\\&=\iota_{\der{E}, D'_1}\circ k'_1\\&=\iota_{\der{E}, K'_0}\\&= \iota_{\der{E}, R_0} \circ r_0
	\end{align*}

@@

\begin{lemma}
Let $(P, \{p_0, p_1\})$ be the pullback of $f_1$ along $g_1$, then there exists $z_0\colon P\mto D'_0$ and $z_1\colon P\to D'_1$ making the following diagram commutative
\[\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){i_0}|(.7)\hole
	\ar[dr]^{h_0}&& L_1\ar@/^.8cm/[ddll]^(.2){i_1}
	\ar[dl]_{m_1}\\&K_0\ar[dr]^{k_0}\ar@{>->}[dl]_{l_0}
	\ar@/_.6cm/[ddrr]|(.36)\hole^(.65){u_0}\ar[ur]^{r_0}&& G_1 &&
	K_1
	\ar@/^.6cm/[ddll]|(.36)\hole_(.65){u_1}\ar[dl]_{k_1}\ar@{>->}[lu]_{l_1}
	\ar[dr]^{r_1}\\L_0
	\ar@/_.6cm/[ddrr]_(.2){i'_1}|(.31)\hole|(.81)\hole
	\ar[dr]^(.4){m_0}|(.61)\hole && D_0
	\ar@{>->}[dl]|(.4)\hole_(.65){f_0}\ar[ur]|(.5)\hole^(.7){g_0}&&D_1
	\ar[dr]|(.4)\hole^(.65){g_1}
	\ar@{>->}[ul]|(.5)\hole_(.65){f_1}&&R_1\ar@/^.6cm/[ddll]^(.2){i'_0}|(.31)\hole|(.81)\hole\ar[dl]_{h_1}|(.6)\hole\\&G_0
	&&P\ar@{.>}[dr]_{z_1}	\ar@{>.>}[dl]^{z_0}\ar[ur]^(.4){p_1}\ar@{>->}[ul]_(.4){p_0}&&G_2\\L_1	\ar@/^.6cm/[uurr]^(.2){i_1} \ar[ur]_(.35){m'_0}|(.61)\hole && D'_0	\ar@{>->}[ul]|(.4)\hole_(.65){f'_0}\ar[dr]|(.5)\hole_(.65){g'_0}	&&D'_1\ar[ur]|(.4)\hole^(.65){g'_1} \ar@{>->}[dl]|(.5)\hole^(.65){f'_1}	&& R_0 \ar[ul]^(.35){h'_1}|(.61)\hole\ar@/_.6cm/[uull]_(.2){i_0}\\ &K_1	\ar[ur]_{k'_0} \ar[dr]_{r_1}	\ar@{>->}[ul]^{l_1}\ar@/^.6cm/[uurr]_(.65){u_1}&&G'_1&& K_0\ar[ur]_{r_0} \ar[ul]^{k'_1} \ar@{>->}[dl]^{l_0} \ar@/_.6cm/[uull]^(.65){u_0}\\&& R_1	\ar[ur]_{h'_0}\ar@/^.8cm/[uurr]^(.2){i'_0}&& L_0 \ar[ul]^{m'_1} \ar@/_.8cm/[uull]_(.2){i'_1} |(.69)\hole }\] 
moreover, the following squares are all pushouts, and the first two are also pullbacks.
\[\xymatrix{P \ar@{>->}[r]^{z_0} \ar@{>->}[d]_{p_0}& D'_0 \ar@{>->}[d]^{f'_0} &  P \ar@{>->}[d]_{z_0} \ar[r]^{z_1} & D'_1 \ar@{>->}[d]^{f'_1} & P\ar[r]^{z_1} \ar[d]_{p_1}&D'_1 \ar[d]^{g'_1}\\ D_0 \ar@{>->}[r]_{f_0} & G_0 & D'_0 \ar[r]_{g'_0}& G'_1 & D_1 \ar[r]_{g_1}& G_2}\]
\end{lemma}
\begin{proof}
	Define $z_0$ and $z_1$ to be, respectively, $t_0\circ v$ and $t_1\circ v$. To prove the commutativity of the diagram in the thesis, first of all we can notice that
	\begin{align*}
		g'_0\circ z_0\\&= g'_0\circ t_0\circ v\\&=f'_1\circ t_1\circ v\\&=f'_1\circ z_1
	\end{align*}
	To prove the other identities, it is enough to compute and use the identities obtained in the previous point. On the one hand we have
	\[	\begin{split}f'_0\circ z_0&=f'_0\circ t_0\circ v\\&=f'_0\circ \phi_{Q_0}\circ q_0\\&=a_0\circ q_0\\&=f_0\circ p_0 
	\end{split} \qquad \begin{split}
		g'_1\circ z_1&=g'_1\circ t_1\circ v\\&=g'_1\circ \phi_{Q_1}\circ q_1\\&=\phi_{G_2}\circ b_1\circ q_1\\\phi_{G_2}\circ g_1\circ p_1\\&=
	\end{split}\]

@@

\begin{lemma}%{lemma}{lemVTame}
	\label{bono}
	All root-preserving graphical rewriting systems are well-switching.
\end{lemma}
